\section{Basis Conversion}\label{sec:conversion}

\textbf{Normal to Power basis}
Suppose $G = \lbrace g_1, \ldots, g_n \rbrace$, $\alpha$ is a normal element of $K$ and $u = \begin{bmatrix}u_1 & \ldots & u_n \end{bmatrix} \in F^n$ is given as the coefficient vector of $u$ in the normal basis.
In order to write $u $ in the power basis one can compute 
\begin{equation}\label{eq:norm2pwrmat}
\left[\begin{array}{c|c|c}
\overline{g_1(\alpha)} & \cdots & \overline{g_n(\alpha)} 
\end{array}\right]\cdot 
\begin{bmatrix}
u_1 & \cdots & u_n
\end{bmatrix}^t,
\end{equation}
where $\overline{g_i(\alpha)}$ is the coefficient vector of $g_i(\alpha)$. The result of \eqref{eq:norm2pwrmat} is the coefficient vector of
\begin{equation}\label{eq:norm2pwrpol}
\sum_{i = 1}^n u_i g_i(\alpha).
\end{equation}
Note that \eqref{eq:norm2pwrmat} shows that conversion from normal to power basis is transpose of the problem of computing the projections of 
conjugates of $\alpha$ which is solved in Section \ref{sec:osum}. Here we present an explicit algorithm to solve this problem i.e. to compute
\eqref{eq:norm2pwrpol} for abelian and metacyclic $G$. Before stating the algorithm for basis conversion, we present the last
ingredient needed. 

\begin{lemma}\label{lem:hornerrep}
Under the Assumption \ref{assum}, suppose $g_1, \ldots g_r \in G$ and $h_{i_1 \cdots i_r} \in K$ for $0 \leq j \leq r, \, 
1 \leq i_j \leq s_j, \, \prod_{j = 1}^r s_j \leq  \lceil \sqrt{\vert G \vert} \rceil $ are given. the elements 
$$g_r^{i_r}\cdots g_1^{i_1}(h_{i_1 \cdots i_r}) 0 \leq j \leq r, \, 
1 \leq i_j \leq s_j, \, \prod_{j = 1}^r s_j \leq  \lceil \sqrt{\vert G \vert} \rceil $$
can be computed using $\thecost$ operations over $F$.
\end{lemma}

\begin{proof}
Let $h_{i_1 \cdots i_r}^{(0 \cdots 0)} = h_{i_1 \cdots i_r}$ and $$g_r^{l_r} \cdots g_1^{l_1}(h_{i_1 \cdots i_r}^{(u_1 \cdots u_r)}) = h_{i_1 \cdots i_r}^{(u_1+l_1 \cdots u_r+l_r)}$$ for $0 \leq j \leq r$ and $0 \leq l_j$.
Assume $0 \leq t < r$ is fixed and $$h_{i_1 \cdots i_r}^{(i_1 \cdots i_{t-1}0 \cdots 0)}, \, 0 \leq j < t, \, 
1 \leq i_j \leq s_j,$$ is already computed. We compute 
$$h_{i_1 \cdots i_r}^{(i_1 \cdots i_{t}0 \cdots 0)}, \, 0 \leq j \leq t, \, 1 \leq i_j \leq s_j,$$
by iteratively applying the following method. In $m$-th iteration we apply $g_t^{2^{m-1}}$ to $h_{i_1 \cdots i_r}^{(i_1 \cdots i_{t}0 \cdots 0)}$ for $j \leq t-1$, $0 \leq i_j \leq s_j$ and 
$$i_t \in \left(\cup_{k = 1}^{\infty} \lbrace (2k-1)2^{m-1}, \ldots , (2k)2^{m-1}-1 \rbrace\right) \cap \lbrace 0 , \ldots , s_t \rbrace $$
At each iteration we have to compute $O(\vert \lceil \sqrt{G} \rceil \vert)$ modular composition which can be done by applying
Lemma \ref{lem:modcom} and $g^{2^i}$ can be computed by applying Lemma
\ref{lem:selfcomp}. This gives the claimed complexity.
\end{proof}

\textbf{Abelian extension.} Under Assumption \ref{assum}, suppose $G$ is given by \eqref{eq:abeliangrp}, $\alpha$ is a normal
element and $u = \begin{bmatrix} u_1 \cdots u_n \end{bmatrix}^t \in F^n$ is an element of $K$. Moreover, let 
$\prod s_i = O(\lceil \sqrt{n} \rceil)$. The basis conversion from normal to power basis is equivalent of computing \eqref{eq:norm2pwrpol} and
the later problem can be solved in three steps. Before presenting the algorithm, note that \eqref{eq:norm2pwrpol} can be written
as 
$$\sum_{\substack{1 \leq j \leq r\\ 0 \leq i_j < e_j}} u_{i_1 \cdots i_r} g_1^{i_1} \cdots g_r^{i_r}(\alpha) = \sum_{\substack{1 \leq j \leq r\\ 0 \leq i_j \leq s_j}} \overline{g_1}^{i_1} \cdots \overline{g_r}^{i_r} (h_{i_1 \cdots i_r}),$$ 
where $$h_{i_1 \cdots i_r} = \sum_{\substack{1 \leq k \leq r\\ 0 \leq j_k < s_k}} u^{(i_1 \cdots i_r)}_{j_1 \cdots j_r}
g_1^{j_1} \cdots g_r^{j_r}(\alpha)$$ and $\overline{g_j} = g_j^{s_j}$.

\textbf{Step 1.} We begin the process by computing $$g_1^{i_1} \cdots g_r^{i_r}(\alpha), \, 1 \leq j \leq r, 0 \leq i_j \leq s_i.$$
Using Lemma \ref{lem:selfcomp}, this can be done using $\osumcosttilde$ operations in $F$.

\textbf{Step 2.} In order to get $h_{i_1 \cdots i_r}$ multiply the matrix with rows $g_1^{i_1} \cdots g_r^{i_r}(\alpha)$ by the
matrix with columns the coefficients of $h_{i_1 \cdots i_r}$. This operation can be done in $O(n^{1/2 \cdot \omega(2)})$.

\textbf{Step 3.} Use Lemma \ref{lem:hornerrep} to compute \eqref{eq:norm2pwrpol}. 

\textbf{Metacyclic extension.} with the same assumptions as in the abelian case, suppose $G$ is given by \eqref{eq:metacyclic}. In
this case \eqref{eq:norm2pwrpol} can be rewritten as
$$\sum_{i = 0}^m \sum_{0}^s u_{ij}\sigma^i \tau^j(\alpha) = \sum_{i = 0}^{\lceil \sqrt{m} \rceil} \overline{\sigma}^i(h_i),$$
where $h_i = \sum_{j = 0}^{\sqrt{m}/\sqrt{s}}\sum_{k =0}^{s} a_{jk}^{(i)}\sigma^j \tau^k (\alpha)$ and 
$\overline{\sigma} = \sigma^{\lceil \sqrt{ms} \rceil}$.
Note that without loss of generality we can assume $s \leq m$. Similar to the abelian case, we can do the computation in three steps.

\textbf{Step 1.} Compute $\sigma^i\tau^j(\alpha)$ for $0 \leq i \leq \lceil \sqrt{m} /\sqrt{s} \rceil$ and $0 \leq j < s$.
Using Lemma \ref{lem:selfcomp}, this can be done using $\osumcosttilde$ operations in $F$.

\textbf{Step 2.} In order to get $h_{i}$ multiply the matrix with rows $\sigma^i\tau^j(\alpha)$ by the
matrix with columns the coefficients of $h_{i}$. This operation can be done in $O((ms)^{1/2 \cdot \omega(2)})$.

\textbf{Step 3.} Use Lemma \ref{lem:hornerrep} to compute \eqref{eq:norm2pwrpol}. 

\textbf{Power Basis to normal basis.}

Now assume $u \in K$ is given in power basis. The goal is to find $c_i$'s in $F$ such that, 
$$u = \sum_{i = 0}^{n-1} c_i g_i(\alpha).$$
Then for any element $g_j$ of $G$ we have
$$g_j(u) = \sum_{i = 0}^{n-1} c_i g_jg_i(\alpha).$$
On the other hand if $l$ is a random projection of $K$ to $F$, we get 
$$l(g_j(u)) = \sum_{i = 0}^{n-1} c_i l(g_jg_i(\alpha)).$$
Hence, in order to find $c_i$'s it is enough to solve the linear system
\begin{equation}\label{eq:convlinsys}
l(M_G(\alpha)) \textbf{x} = \begin{bmatrix} l(g_i(u)) \end{bmatrix}_{0 \leq i <n}
\end{equation}
Hence the conversion problem can be solved in 2 steps 

\textbf{Step 1.} compute $\begin{bmatrix} l(g_i(u)) \end{bmatrix}_{0 \leq i <n}$ which can be done using
the algorithms of Section \ref{sec:osum}. This can be carried out using $\thecost$ operations in $F$.

\textbf{Step 2.} compute $l(M_G(\alpha))$ and then solving the linear system \eqref{eq:convlinsys}.

\textbf{Abelian extension.}
It is already mentioned that multiplying by $l(M_G(\alpha))$ is equivalent of multiplication in a multivariate ring, by 
$s_{\alpha, \ell}$. Hence by considering the group algebra version of \eqref{eq:convlinsys} we have to solve $$s_{\alpha,\ell} x 
= s_u$$ for $x$ 
in $F[G]$. This can be done by using the tools provided in Section ?. Computing the inverse of $s_{\alpha,\ell}$ costs the same 
as testing its invertiblity. Having the inverse we get the solution with a single multiplication in $F[G]$.

\textbf{Metacyclic extension.}

In order to solve \eqref{eq:convlinsys}, in the metacyclic case, we start with understanding the structure of $M_G(\alpha)$. 
Assume the first row of $M_G(\alpha)$ is 
$$\begin{bmatrix}
\sigma^0 \tau^0 &  \cdots & \sigma^{n-1} \tau^0 & \sigma^0 \tau &  \cdots & \sigma^{n-1} \tau & \sigma^0 \tau^s &  \cdots & 
\sigma^{n-1} \tau^s 
\end{bmatrix}$$
Now by considering the actions (respectively) of
$$\tau^0 \sigma^1, \ldots , \tau^0 \sigma^{n-1}, \tau^1 \sigma^0, \ldots , \tau^1 \sigma^{n-1} , \cdots \tau^s \sigma^0, \ldots 
, \tau^s \sigma^{n-1}$$
on the first row we get $M_G(\alpha)$.
Note that here we are using both presentation of elements in $G$ as $\sigma^i\tau^j$ and $\tau^j\sigma^i$. Using above argument 
it is not hard to see $M_G(\alpha)$ as a block matrix with blocks generated by the action of $\tau^l \sigma^1, \ldots , \tau^l 
\sigma^{n-1}$ on $\sigma^0 \tau^k ,  \ldots , \sigma^{n-1} \tau^k$. In other word the $lk$-th block of $M$ is 
$\left( \tau^l \sigma^{(i-1)+(j-1)} \tau^k \right)_{ij}$ for $1 \leq i,j \leq n$. This shows that each block is a Hankel $n \times n$ matrix.



%%% Local Variables:
%%% mode: latex
%%% TeX-master: "NormalBasisCharZero"
%%% End:
